\pkg{depmixS4} : An \proglang{R}-package for hidden Markov models

### 2009-08-05 papers/jss/article.tex
Minor changes to jss paper, bib update.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@306 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Notation issues ...
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@307 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%% an abstract and keywords
\Abstract{
	\pkg{depmixS4} implements a general framework for defining and estimating
	 dependent mixture models in the \proglang{R} programming language
	\citep{R2009}. This includes standard Markov models, latent/hidden Markov
	models, and latent class and finite mixture distribution models. The models 
	can be fitted on mixed multivariate data with distributions from the 
	\code{glm} family, the logistic multinomial, or the multivariate normal 
	distribution. Other distributions can be added easily, and an example is 
	provided. Parameter are estimated by the EM algorithm or, when (linear) 
	constraints are imposed on the parameters, by direct numerical optimization 
	with the \pkg{Rdonlp2} routine.
}

@@

@@

Markov and latent Markov models are frequently used in the social
sciences, in different areas and applications.  In psychology, they
are used for modelling learning processes, see \citet{Wickens1982},
for an overview, and \citet{Schmittmann2006} for a recent application.
In economics, common latent Markov models are so-called regime
switching models (see e.g., \citealp{Kim1994} and \citealp{Ghysels1994}).
Further applications include speech recognition \citep{Rabiner1989},
EEG analysis \citep{Rainer2000}, and genetics \citep{Krogh1998}.  In
those latter areas of application, latent Markov models are usually
referred to as hidden Markov models. For more examples of applications, 
see e.g.\ \citet[][chapter~1]{Cappe2005}.

@@

	\item to be easily extensible, in particular, to allow users to
  easily add new uni- or multivariate response distributions, and similarly 
  add of other transition models, e.g., continuous time observation
	models.

@@

The data considered here have the general form $O_{1}^{1}, \ldots,
O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$, \ldots, $O_{T}^{1},
\ldots, O_{T}^{m}$ for an $m$-variate time series of length $T$.  As
an example, consider a time series of responses generated by a single
participant in a response time experiment. The data consists of three
variables, response time, response accuracy, and a covariate which is a pay-off
variable reflecting the relative reward for speeded and/or accuracy responding. 
These variables are measured on 168, 134 and 137 occasions respectively (in
Figure~\ref{fig:speed} the first part of this series is plotted). These data
are more fully described in \citet{Dutilh2009}.

@@

The latent Markov model is commonly associated with data of this type,
although usually only multinomial response variables are considered.  However, 
common estimation procedures, such as those implemented in
\citet{Pol1996}, are not suitable for long time series due to underflow
problems.  In contrast, the hidden Markov model is typically only used
for `long' univariate time series 
\citep[][, chapter~1]{Cappe2005}.  
In the next sections, the
likelihood equation and estimation procedures for the dependent mixture model 
are described for data of the above form. We use the term
``dependent mixture model'' because one of the authors (Ingmar Visser)
thought it was time for a new name for these models\footnote{Only
later did I find out that \citet{Leroux1992} already coined the term
dependent mixture models in an application with hidden Markov mixtures
of Poisson count data.}.

The dependent mixture model is defined by the following elements:
\begin{enumerate}

	\item a set $\mathcal{S}$ of latent classes or states $S_{i},\, i=1,
	\ldots , n$,

	%\item a set of matrices $\mat{A}_t$ of transition probabilities $a_{ijt} = P_t(S_j|S_i)$ 
	%for the transition from state $S_{i}$ to state $S_{j}$ at time $t$,

	\item a set $\mathcal{A}$ of transition probability functions $a_{ij}(\vc{x}_t) = 
	Pr(S_t=j|S_{t-1}=i,\vc{x}_t)$, giving the probability of a transition from state 
	$i$ to state $j$ with covariate $\vc{x}_t$,

	\item a set $\mathcal{B}$ of observation density functions $b_j^k(\vc{x}_t) = p(O_{t}^k|S_t = j, \vc{x}_t)$ that
	provide the conditional densities of observations $O_{t}^k$ 
	associated with latent class/state $j$ and covariate $\vc{x}_t$,

	\item a set $\mathcal{P}$ of latent class/state initial probability functions
	$\pi_{i}(\vc{x}_1) = Pr(S_1 = i|\vc{x}_1)$, giving the probability of class/state
	$i$ at time $t=1$ with covariate $\vc{x}_1$.

The dependent mixture model is defined by the following equations: 
%\begin{align}
%	S_{t} &= A S_{t-1}, t=2, \ldots, T \\
%	O_{t} &= b(O_{t}|S_{t}), 
%\end{align}
\begin{align}
	Pr(S_{t} = j) &= \sum_{i=1}^n a_{ij}(\vc{x}_t) Pr(S_{t-1} = i), t=2, \ldots, T \\
	p(O_{t}^k|S_t &= j, \vc{x}_t) = b_j^k(\vc{x}_t)
\end{align}
%where $S_{t}$ is a sequence of hidden states, $A$ is a transition
%matrix, $O_{t}$ is an (possibly multivariate) observation and $b$ is a
%density function for $O_{t}$ conditional on the hidden state $S_{t}$.
In the example data above, $b_j^k$ could be a Gaussian distribution
function for the response time variable, and a Bernoulli for the
accuracy data.  In the models we are considering here, both the
transition probabilities $\mat{A}$ and the initial state probabilities $\pi$
may depend on covariates as well as the response distributions $\vc{B}$. 
See for example \citet{Fruhwirth2006} for an overview of 
hidden Markov models with extensions.

@@

The log-likelihood of hidden Markov models is usually computed by the
so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or
rather by the forward part of this algorithm.  \cite{Lystig2002}
changed the forward algorithm in such a way as to allow computing the
gradients of the log-likelihood at the same time.  They start by
rewriting the likelihood as follows (for ease of exposition the
dependence on the model parameters is dropped here):
\begin{equation}
	L_{T} = Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) = \prod_{t=1}^{T} 
Pr(\vc{O}_{t}|\vc{O}_{1}, 
	\ldots, \vc{O}_{t-1}), 
	\label{condLike}
\end{equation}
where $Pr(\vc{O}_{1}|\vc{O}_{0}):=Pr(\vc{O}_{1})$. Note that for a 
simple, i.e.\ observed, Markov chain these probabilities reduce to 
$Pr(\vc{O}_{t}|\vc{O}_{1},\ldots, 
\vc{O}_{t-1})=Pr(\vc{O}_{t}|\vc{O}_{t-1})$.
The log-likelihood can now be expressed as:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log[Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, 
\vc{O}_{t-1})].
	\label{eq:condLogl}
\end{equation}

To compute the log-likelihood, \cite{Lystig2002} define the following 
(forward) recursion:
\begin{align}
	\phi_{1}(j) &:= Pr(\vc{O}_{1}, S_{1}=j) = \pi_{j} b_{j}(\vc{O}_{1}) 
	\label{eq:fwd1} \\
\begin{split}
	\phi_{t}(j) &:= Pr(\vc{O}_{t}, S_{t}=j|\vc{O}_{1}, \ldots, 
\vc{O}_{t-1}) \\
	&= \sum_{i=1}^{N} [\phi_{t-1}(i)a_{ij}b_{j}(\vc{O}_{t})] \times 
(\Phi_{t-1})^{-1},
	\label{eq:fwdt} 
\end{split} 
\end{align}
where $\Phi_{t}=\sum_{i=1}^{N} \phi_{t}(i)$. Combining 
$\Phi_{t}=Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1})$, and 
equation~(\ref{eq:condLogl}) gives the following expression for the 
log-likelihood:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log \Phi_{t}.
	\label{eq:logl}
\end{equation}

>>>>

%% an abstract and keywords
\Abstract{

	\pkg{depmixS4} implements a general framework for defining and
	estimating dependent mixture models in the \proglang{R}
	programming language \citep{R2009}.  This includes standard Markov
	models, latent/hidden Markov models, and latent class and finite
	mixture distribution models.  The models can be fitted on mixed
	multivariate data with distributions from the \code{glm} family,
	the logistic multinomial, or the multivariate normal distribution.
	Other distributions can be added easily, and an example is
	provided with the exgaus distribution.  Parameter are estimated by
	the EM algorithm or, when (linear) constraints are imposed on the
	parameters, by direct numerical optimization with the
	\pkg{Rdonlp2} routine.  }

@@

\begin{center}
\bf{DRAFT: DO NOT QUOTE WITHOUT CONTACTING AUTHOR}
\end{center}

@@

Markov and latent Markov models are frequently used in the social
sciences, in different areas and applications.  In psychology, they
are used for modelling learning processes, see \citet{Wickens1982},
for an overview, and e.g.\ \citet{Schmittmann2006} for a recent
application.  In economics, latent Markov models are so-called regime
switching models (see e.g., \citealp{Kim1994} and
\citealp{Ghysels1994}).  Further applications include speech
recognition \citep{Rabiner1989}, EEG analysis \citep{Rainer2000}, and
genetics \citep{Krogh1998}.  In those latter areas of application,
latent Markov models are usually referred to as hidden Markov models.
For more examples of applications, see e.g.\
\citet[][chapter~1]{Cappe2005}.

@@

	\item to be easily extensible, in particular, to allow users to
	easily add new uni- or multivariate response distributions, and
	similarly for the addition of other transition models, e.g.,
	continuous time observation models

@@

The data considered here have the general form $\vc{O}_{1:T}=O_{1}^{1}, \ldots,
O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$, \ldots, $O_{T}^{1},
\ldots, O_{T}^{m}$ for an $m$-variate time series of length $T$.  As
an example, consider a time series of responses generated by a single
participant in a response time experiment. The data consists of three
variables, response time, response accuracy, and a covariate which is a pay-off
variable reflecting the relative reward for speeded and/or accuracy responding. 
These variables are measured on 168, 134 and 137 occasions respectively (in
Figure~\ref{fig:speed} the first part of this series is plotted). These data
are more fully described in \citet{Dutilh2009}.

@@

The latent Markov model is commonly associated with data of this type,
although usually only multinomial response variables are considered.
However, common estimation procedures, such as those implemented in
\citet{Pol1996}, are not suitable for long time series due to
underflow problems.  In contrast, the hidden Markov model is typically
only used for `long' univariate time series \citep[][,
chapter~1]{Cappe2005}.  In the next paragraphs, the likelihood
equation and estimation procedures for the dependent mixture model are
described for data of the above form.  We use the term ``dependent
mixture model'' because one of the authors (Ingmar Visser) thought it
was time for a new name for these models\footnote{Only later did I
find out that \citet{Leroux1992} already coined the term dependent
mixture models in an application with hidden Markov mixtures of
Poisson count data.}.

The likelihood of the dependent mixture model conditional on the
unknown (or hidden) state sequence $S_{1:T}$ and the model is given
by:
\begin{equation}
	\Prob( \vc{O}_{1:T} | \vc{S}_{1:T}, \greekv{\theta} ) = 
	\prod_{t=1}^{T}  \Prob( \vc{O}_{t} | S_{t},  \greekv{\theta}),
\end{equation}
where $\greekv{\theta}$ is the parameter vector of the model.  To arrive
at the likelihood of the data given the parameter vector
$\greekv{\theta}$, i.e.\ without the state sequence, we need to sum
above likelihood over all possible state sequences. This likelihood 
is written as: 
\begin{equation}
	\Prob(\vc{O}_{1:T}|\vc{S}_{1:T},\greekv{\theta}) 
	\sum_{all Ss} \pi_{i}(\vc{z}_1) \vc{b}_{1}(\vc{O}_{1}) 
	\prod_{t=1}^{T-1} a_{ij}(z_{t}) \vc{b}_{j}(\vc{O}_{t+1}|\vc{z}_{t+1}),
\end{equation}
where we have the following elements:
\begin{enumerate}

	\item $S_{t}$ is an element of $\mathcal{S}=\{1\ldots n\}$, a set
	of $n$ latent classes or states; we write for short $S_{i}$ to
	denote $S_{t}=i$.

	\item $\pi_{i}(\vc{z}_1) = \Prob(S_1 = i|\vc{z}_1)$,
	giving the probability of class/state $i$ at time $t=1$ with
	covariate $\vc{z}_1$.

	\item $a_{ij}(\vc{z}_t) = \Prob(S_{t+1}=j|S_{t}=i,\vc{z}_t)$,
	provides the probability of a transition from state $i$ to state
	$j$ with covariate $\vc{z}_t$,

	\item $\vc{b}_{j}$ is a vector of observation densities
	$b_j^k(\vc{z}_t) = \Prob(O_{t}^k|S_t = j, \vc{z}_t)$ that provide the
	conditional densities of observations $O_{t}^k$ associated with
	latent class/state $j$ and covariate $\vc{z}_t$, $j=1, \ldots, n$,
	$k=1, \ldots, m$. 
\end{enumerate}

% The dependent mixture model is defined by the following equations: 
% %\begin{align}
% %	S_{t} &= A S_{t-1}, t=2, \ldots, T \\
% %	O_{t} &= b(O_{t}|S_{t}), 
% %\end{align}
% \begin{align}
% 	Pr(S_{t} = j) &= \sum_{i=1}^n a_{ij}(\vc{x}_t) Pr(S_{t-1} = i), t=2, \ldots, T \\
% 	p(O_{t}^k|S_t &= j, \vc{x}_t) = b_j^k(\vc{x}_t)
% \end{align}
%where $S_{t}$ is a sequence of hidden states, $A$ is a transition
%matrix, $O_{t}$ is an (possibly multivariate) observation and $b$ is a
%density function for $O_{t}$ conditional on the hidden state $S_{t}$.

In the example data above, $b_j^k$ could be a Gaussian distribution
function for the response time variable, and a Bernoulli for the
accuracy data.  In the models we are considering here, both the
transition probability functions $a_{ij}$and the initial state
probability functions $\greekv{\pi}$ may depend on covariates as well
as the response distributions $\vc{b_{j}^{k}}$.  See for example
\citet{Fruhwirth2006} for an overview of hidden Markov models with
extensions.

@@

The log-likelihood of hidden Markov models is usually computed by the
so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or
rather by the forward part of this algorithm.  \cite{Lystig2002}
changed the forward algorithm in such a way as to allow computing the
gradients of the log-likelihood at the same time.  They start by
rewriting the likelihood as follows (for ease of exposition the
dependence on the model parameters is dropped here):
\begin{equation}
	L_{T} = Pr(\vc{O}_{1:T}) = \prod_{t=1}^{T} 
	Pr(\vc{O}_{t}|\vc{O}_{1:t-1}), 
	\label{condLike}
\end{equation}
where $Pr(\vc{O}_{1}|\vc{O}_{0}):=Pr(\vc{O}_{1})$. Note that for a 
simple, i.e.\ observed, Markov chain these probabilities reduce to 
$Pr(\vc{O}_{t}|\vc{O}_{1},\ldots, 
\vc{O}_{t-1})=Pr(\vc{O}_{t}|\vc{O}_{t-1})$.
The log-likelihood can now be expressed as:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log[\Prob(\vc{O}_{t}|\vc{O}_{1}, \ldots, 
\vc{O}_{t-1})].
	\label{eq:condLogl}
\end{equation}

To compute the log-likelihood, \cite{Lystig2002} define the following 
(forward) recursion:
\begin{align}
	\phi_{1}(j) &:= \Prob(\vc{O}_{1}, S_{1}=j) = \pi_{j} b_{j}(\vc{O}_{1}) 
	\label{eq:fwd1} \\
\begin{split}
	\phi_{t}(j) &:= \Prob(\vc{O}_{t}, S_{t}=j|\vc{O}_{1}, \ldots, 
\vc{O}_{t-1}) \\
	&= \sum_{i=1}^{N} [\phi_{t-1}(i)a_{ij}b_{j}(\vc{O}_{t})] \times 
(\Phi_{t-1})^{-1},
	\label{eq:fwdt} 
\end{split} 
\end{align}
where $\Phi_{t}=\sum_{i=1}^{N} \phi_{t}(i)$. Combining 
$\Phi_{t}=\Prob(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1})$, and 
equation~(\ref{eq:condLogl}) gives the following expression for the 
log-likelihood:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log \Phi_{t}.
	\label{eq:logl}
\end{equation}
